---
title: "nlmixr2sir events for loggers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nlmixr2sir events for loggers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_bus <- exists("rxEventListen", asNamespace("rxode2"))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_bus
)
```

```{r no-bus, echo = FALSE, eval = !has_bus, results = "asis"}
cat(
  "> **Note:** the installed rxode2 has no event bus",
  "(`rxode2::rxEventListen()` is missing), so the code in this vignette",
  "is shown but not run.  `runSIR()` works exactly as before; it simply",
  "emits nothing.\n"
)
```

## The rxode2 event bus

rxode2 provides a small in-process event bus.  Operations such as a fit, a
simulation or a post-fit result call `rxode2::rxEventEmit()`, and any function
registered with `rxode2::rxEventListen()` is told about it.  Listeners are
called as `fun(event, ...)`: the event name followed by the payload fields as
named arguments.  A listener that errors gives a warning and never stops the
caller.  See `?rxode2::rxEventListen` for the full API.

Events are delivered only from the outermost operation (scope depth 0).  Work
done inside another operation's scope (`rxode2::rxEventScope()`, or between
`.rxEventEnter()` and `.rxEventExit()`) is silent, so a logger sees one event
for what the user asked for, not one per internal fit or solve.  The depth is
mirrored in the `RXODE2_EVENT_DEPTH` environment variable, so parallel workers
started inside a scope stay silent too.

nlmixr2sir only uses the bus when the installed rxode2 has one; with an older
rxode2, or with no listener registered, nothing changes.

## What `runSIR()` emits

`runSIR()` emits exactly one event: a `fitResult` when it returns, attaching
the SIR result to the fit it was run on.

| Field    | Value |
|----------|-------|
| `fit`    | The `fit` argument as passed to `runSIR()` (captured on entry) |
| `result` | A slimmed copy of the returned `nlmixr2SIR` object (see below) |
| `kind`   | `"sir"` |
| `call`   | The `runSIR()` call, normalized by rxode2 (head `runSIR`; large inlined values replaced by `` `<value>` ``) |

The `result` payload is the SIR summary table with only these attributes kept:
`iterationSummary`, `covMatrix`, `corMatrix`, `sdCorMatrix`, `outputDir`,
`fitName`, `seed` and `fingerprint` (plus `class`, `names` and `row.names`).
The resampled parameter vectors (`resampledMat`) and the raw per-sample
results (`rawResults`) are dropped so a logger stores a small object; when a
`directory` is given they remain in the SIR run directory, whose path is in
`attr(result, "outputDir")`.
The object returned to the user is untouched.

## What is not emitted

- **The internal objective evaluations.**  `runSIR()` enters the bus scope on
  entry, so the many fixed-parameter evaluations it runs (and any
  `fitComplete`, `solveComplete` or `fitUpdate` events they would otherwise
  raise in nlmixr2est or rxode2) are silent.  Parallel workers started through
  `nlmixr2utils` enter the scope themselves.
- **Failed runs.**  When `runSIR()` stops with an error there is no result,
  so it leaves the scope without emitting anything.
- **Non-fits.**  A `fitResult` is only emitted when `fit` is an nlmixr2 fit
  (`nlmixr2FitCore`).  If the summary cannot be built, the scope is left
  silently rather than failing the run.
- **Nested runs.**  When `runSIR()` itself is called inside another operation
  (for example from a driver that is already in a scope), its `fitResult` is
  dropped like any other nested event; the outer operation reports.

## Listening to `runSIR()`

A listener is any `function(event, ...)`.  This one records each event name and
its payload.

```{r listen}
library(nlmixr2sir)

events <- new.env()
events$log <- list()
rxode2::rxEventListen("demo", function(event, ...) {
  events$log[[length(events$log) + 1L]] <- list(event = event, payload = list(...))
})
rxode2::rxEventListeners()
```

```{r fit, message = FALSE, results = "hide"}
three_eta_one_cmt <- function() {
  ini({
    tka <- 0.45
    tcl <- 1.00
    tv <- 3.45
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}

fit <- rxode2::rxEventScope(nlmixr2utils::nlmixr2(
  three_eta_one_cmt,
  data    = nlmixr2data::theo_sd,
  est     = "focei",
  control = list(print = 0L)
))
```

Wrapping the fit in `rxode2::rxEventScope()` keeps it silent, so only the SIR
event is recorded below.  A tiny SIR schedule keeps the example quick (far too
small for real use, so its effective-sample-size warnings are hidden here):

```{r sir, message = FALSE, warning = FALSE, results = "hide"}
set.seed(20261009)
sir <- runSIR(
  fit,
  nSamples  = c(16L, 16L),
  nResample = c(8L, 8L),
  directory = file.path(tempdir(), "sir_events_vignette"),
  control   = runSIRControl(workers = 1L, recover = FALSE)
)
```

All internal evaluations were silent; the listener saw one event:

```{r inspect}
vapply(events$log, `[[`, "", "event")
p <- events$log[[1]]$payload
names(p)
p$kind
p$call
identical(p$fit, fit)
names(attributes(p$result))
is.null(attr(p$result, "resampledMat"))  # dropped from the event ...
is.null(attr(sir, "resampledMat"))       # ... but kept in the returned object
attr(p$result, "outputDir")
```

Remove the listener when done:

```{r unlisten}
rxode2::rxEventUnlisten("demo")
rxode2::rxEventDepth()
```

```{r cleanup, include = FALSE}
unlink(file.path(tempdir(), "sir_events_vignette"), recursive = TRUE)
```

## Full logging with nlmixr2log

The [nlmixr2log](https://github.com/nlmixr2/nlmixr2log) package is the main
consumer of these events.  It listens to the bus and records fits, results
and saved objects in a run log; with it, a `runSIR()` call is stored as one
SIR result attached to its fit rather than as a run for every internal
evaluation.  Any other package can listen in the same way as the example
above.
