---
title: "Sampling Importance Resampling with runSIR()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sampling Importance Resampling with runSIR()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

Standard uncertainty estimates for non-linear mixed-effects model parameters rely on the
covariance matrix produced by the covariance step (an approximation to the inverse
Fisher information matrix).  This approximation performs well when the likelihood
surface is nearly quadratic near the maximum — a condition that often fails for
variance components, highly correlated parameters, or models with a restricted
parameter space.

Sampling Importance Resampling (SIR; Dosne et al. 2013) is a simulation-based
alternative that samples directly from the parameter uncertainty distribution.  The
algorithm:

1. **Samples** parameter vectors from a proposal distribution centred on the MLE and
   using the asymptotic covariance as its shape.
2. **Evaluates** each sampled vector by re-running the model with those parameters
   fixed (one OFV evaluation per sample, equivalent to NONMEM `MAXEVAL=0`).
3. **Weights** each sample by the ratio of the target density (the likelihood) to the
   proposal density (importance weighting).
4. **Resamples** without replacement according to those weights.
5. **Updates** the proposal covariance from the empirical covariance of the resampled
   vectors and repeats for the next iteration.

After convergence the resampled vectors represent draws from the parameter uncertainty
distribution, and empirical percentiles give non-parametric confidence intervals that
require no asymptotic assumption.

`runSIR()` implements this algorithm for `nlmixr2` FOCE-I fits, following the PsN
`sir` workflow where practical. To make SIR the covariance a fit reports, see
`vignette("setCov", package = "nlmixr2sir")`.

---

## Prerequisites

`runSIR()` uses `fit$cov` for the initial proposal when it is available, which is
the usual case and the one this vignette follows.  A successful covariance step is
*not* required: if your fit has none, see
[Running without a covariance step](#running-without-a-covariance-step) below.

SIR also requires at least two random effects (ETAs) so that a multivariate proposal
can be constructed for the omega block.

```{r packages, message = FALSE}
library(nlmixr2data)
library(nlmixr2sir)
```

---

## Example: theophylline single-dose PK

We use `nlmixr2data::theo_sd` — 12 subjects after a single oral dose — with a
one-compartment model that includes IIV on absorption, clearance, and volume.

### 1. Fit the base model

The covariance step runs automatically with FOCE-I.  Request NPDE and CWRES residuals
via `table` for later diagnostics if needed.

```{r fit, message = FALSE, results = "hide"}
three_eta_one_cmt <- function() {
  ini({
    tka <- 0.45 # log absorption rate constant (1/h)
    tcl <- 1.00 # log clearance (L/h)
    tv <- 3.45 # log central volume (L)
    eta.ka ~ 0.6 # IIV on absorption
    eta.cl ~ 0.3 # IIV on clearance
    eta.v ~ 0.1 # IIV on volume
    add.sd <- 0.7 # additive residual error (mg/L)
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}

fit <- nlmixr2utils::nlmixr2(
  three_eta_one_cmt,
  data    = nlmixr2data::theo_sd,
  est     = "focei",
  control = list(print = 0L),
  table   = list(npde = TRUE, cwres = TRUE)
)
```

```{r fit-summary}
fit$parFixedDf
round(as.matrix(fit$omega), 4)
```

---

### 2. Run SIR

The most important arguments are:

| Argument | Default | Meaning |
|---|---|---|
| `nSamples` | `c(1000, 1000, 1000, 2000, 2000)` | Samples per iteration |
| `nResample` | `c(200, 400, 500, 1000, 1000)` | Resamples per iteration |
| `thetaInflation` | `1` | Variance multiplier for theta proposal |
| `omegaInflation` | `1` | Variance multiplier for omega proposal |
| `capCorrelation` | `0.8` | Cap on absolute pairwise correlations |
| `capResampling` | `1` | Max times one sample may be redrawn |
| `recenter` | `TRUE` | Shift proposal mean to best sample when dOFV < 0 |
| `boxcox` | `TRUE` | Use Box-Cox transformed empirical covariance |
| `workers` | `NULL` | Sequential; set > 1 for parallel OFV evaluation |
| `directory` | `NULL` | Output folder (auto-named if `NULL`) |
| `recover` | `TRUE` | Resume from saved state if found |

The example below uses a two-iteration schedule with small sample counts for
illustration.  In a real analysis, use the PsN-recommended defaults (1 000 / 2 000
samples, 200 / 1 000 resamples) or increase counts until the resampled covariance
stabilises.

```{r sir, message = FALSE}
set.seed(20260420)

sir_result <- runSIR(
  fit,
  nSamples = c(16L, 16L),
  nResample = c(8L, 8L),
  directory = file.path(
    tempdir(), paste0("sir_vignette_", format(Sys.time(), "%Y%m%d_%H%M%S"))
  ),
  control = runSIRControl(
    thetaInflation = 1,
    omegaInflation = 1,
    sigmaInflation = 1,
    capCorrelation = 0.8,
    capResampling = 1,
    recenter = TRUE,
    boxcox = TRUE,
    workers = 1L,
    recover = FALSE
  )
)
```

---

### 3. Inspect the results

`print()` shows the final parameter uncertainty table and per-iteration diagnostics:

```{r print-sir}
print(sir_result)
```

Key columns in the summary:

| Column | Meaning |
|---|---|
| `estimate` | MLE from the original fit |
| `sd` | Empirical SD of the resampled vectors |
| `rse` | 100 × sd / \|estimate\| |
| `p2.5`, `p97.5` | 2.5th and 97.5th empirical percentiles (95 % CI) |
| `p50` | Median of the resampled distribution |

The iteration diagnostics report how many samples were drawn, how many OFV evaluations
succeeded, and how many vectors were resampled.  A healthy run has few failures
(`nFailed` ≈ 0) and no negative dOFV values (`nNegativeDOFV` = 0; if present, the
proposal mean is recentered automatically when `recenter = TRUE`).

---

### 4. Re-compute the summary from the resampled matrix

`sirSummary()` recomputes the table from the raw resampled matrix, which is also
useful when you want to summarise a subset of iterations or apply the function to a
matrix from an external source.

```{r sir-summary}
final_mat <- attr(sir_result, "resampledMat")
sir_summ <- sirSummary(final_mat, fit)
print(sir_summ)
```

Empirical covariance and correlation matrices between parameters are available as
attributes:

```{r cov-cor}
round(attr(sir_summ, "covMatrix"), 6)
round(attr(sir_summ, "corMatrix"), 4)
```

---

### 5. Diagnostic plots

`plot()` returns a `ggplot2` object.  Six views are available; the convergence
plot is the one to read first.

#### Parameter distributions

Histograms of the final resampled vectors with the MLE marked in orange.  The shape
reveals whether the posterior is symmetric and whether any parameter is near a bound.

```{r plot-params, fig.cap = "Final resampled parameter distributions"}
plot(sir_result, type = "parameters", bins = 12L)
```

#### dOFV distributions

Histograms of ΔOFV = OFV(sample) − OFV(fit) per iteration.  A good proposal yields a
distribution concentrated near 0–10; a very flat or right-skewed dOFV histogram
suggests the proposal is too narrow or too wide.

```{r plot-dofv, fig.cap = "dOFV distributions by iteration"}
plot(sir_result, type = "dofv", bins = 12L)
```

#### Resampling probabilities

Bar chart of resampling probabilities per sample per iteration.  A healthy run has
many samples with non-negligible probability; a run dominated by a few high-weight
samples indicates a poor proposal (too wide).

```{r plot-resample, fig.cap = "Resampling probabilities by iteration"}
plot(sir_result, type = "resampling")
```

#### Convergence: dOFV against the reference chi-square

The primary SIR diagnostic.  For each iteration it draws the empirical ΔOFV quantile
curve for the proposal and for the SIR posterior, against a reference chi-square on
the number of estimated parameters.  **Convergence reads as the SIR curve settling
onto the reference.**

If the first iteration's proposal curve falls *below* the reference for more than a
quarter of the quantiles, the proposal is too narrow.  Resampling cannot recover from
that — the samples needed simply were not drawn — so `runSIR()` warns and recommends
restarting with inflation.

```{r plot-convergence, fig.cap = "dOFV against the reference chi-square"}
plot(sir_result, type = "convergence", nReplicate = 100L)
```

#### Confidence intervals by iteration

Proposal and SIR intervals per parameter per iteration.  Use it to judge whether the
uncertainty has stopped moving: if the last two iterations differ noticeably, the run
has not settled.

```{r plot-intervals, fig.cap = "Interval per parameter, by iteration"}
plot(sir_result, type = "intervals")
```

#### RSE and correlation

RSE% on the diagonal, correlations off it.  The diagonal is annotated by the
confidence-interval asymmetry ratio `(high - median) / (median - low)`.  A symmetric
normal-approximation covariance reports a single standard error per parameter and
cannot express that asymmetry — showing it is a large part of why SIR is run.

```{r plot-rsecor, fig.cap = "RSE and correlation, with CI asymmetry"}
plot(sir_result, type = "rsecor")
```

---

### 6. Output files

`runSIR()` writes several files to the output directory:

```{r list-files}
list.files(attr(sir_result, "outputDir"))
```

| File | Contents |
|---|---|
| `raw_results.csv` | Canonical shared raw-results table with the reference fit and final resampled vectors |
| `raw_results.rds` | RDS copy of the canonical raw-results table |
| `raw_results_header.json` | Schema metadata for the canonical raw-results table |
| `summary_iterations.csv` | Per-iteration diagnostic summary |
| `sample_rejection_summary.txt` | Counts of rejected samples (bounds, omega non-PD) |
| `sir_results.csv` | Final parameter summary (same as the printed table) |
| `sir_seed.rds` | Persisted SIR master seed |
| `sir_state.rds` | Full internal state for recovery or adding iterations |

---

### 7. Advanced use

#### Resuming an interrupted run

By default `recover = TRUE` so that if a long SIR run is interrupted, re-calling
`runSIR()` with the same `directory` will pick up from the last completed iteration.

#### Adding iterations

To append more iterations to a completed run, use `addIterations = TRUE` with a new
sample schedule.  This is useful when the last two iterations still show noticeable
changes in the resampled covariance and you want to extend an existing run rather than
start over.

#### Parallel OFV evaluation

OFV evaluations are the bottleneck.  Setting `workers > 1` runs them in parallel using
the `future` / `future.apply` packages.  This is most helpful for larger production
SIR schedules, where many OFV evaluations can be spread across workers.

Sequential evaluation (`workers = NULL` or `workers = 1`) does not require `future` or
`future.apply`.

Whenever `workers > 1`, `workers * rxThreads` must not exceed the machine's core
count — each worker is a separate process running its own rxode2 thread pool, so
oversubscribing slows the run down rather than speeding it up.

```{r parallel, eval = FALSE}
runSIR(
  fit,
  control = runSIRControl(workers = 4, rxThreads = 2)
)
```

---

## Running without a covariance step

The case SIR is most wanted for is the one where the covariance step failed — and
that is precisely the case where an asymptotic standard error is unavailable. A
covariance step is not required.

Suppose the fit has no `$cov`:

```{r nocov-fit, message = FALSE, results = "hide"}
fit_nocov <- nlmixr2utils::nlmixr2(
  three_eta_one_cmt,
  data = nlmixr2data::theo_sd,
  est = "focei",
  control = list(print = 0L, covMethod = "")
)
```

```{r nocov-check}
is.null(fit_nocov$cov)
```

There are three ways to give SIR a proposal anyway.

### From relative standard errors

State the uncertainty directly as RSE percentages. A scalar `rseTheta` is applied to
OMEGA and residual error too; pass `rseOmega` and `rseSigma` to differentiate them,
or pass a vector with one value per estimated element of a class.

```{r nocov-rse, eval = FALSE}
runSIR(
  fit_nocov,
  nSamples = c(16L, 16L),
  nResample = c(8L, 8L),
  control = runSIRControl(rseTheta = 30)
)
```

Diagonal variances come out as `(rse * estimate / 100)^2`. OMEGA off-diagonals are
derived from the two diagonals they connect. Inflation cannot be combined with this
route — the RSE already states the width, so widening it on top would make the
stated RSE a fiction; widen the RSE instead.

### From a supplied covariance matrix

Pass a matrix, a NONMEM-style `.cov` file, or the string `"identity"`. A named matrix
may cover only some parameters; the rest are filled with zero and then made positive
definite.

```{r nocov-covmat, eval = FALSE}
runSIR(
  fit_nocov,
  control = runSIRControl(covmatInput = "identity", thetaInflation = 0.05)
)
```

`"identity"` together with inflation is the cheap "any diagonal proposal" route: the
identity is almost certainly the wrong scale, so the inflation factor is doing the
real work of setting the width.

### From a raw-results file

Seed the first proposal from parameter vectors that already exist — from an earlier
SIR run, or from a bootstrap. The empirical mean and covariance of those vectors
become the iteration-1 proposal, which is what PsN calls iteration 0.

```{r nocov-rawres, eval = FALSE}
runSIR(
  fit_nocov,
  control = runSIRControl(
    rawresInput = "path/to/raw_results.csv",
    offsetRawres = 1L,
    inFilter = function(d) d$objf < 1000
  )
)
```

`offsetRawres` skips low-numbered samples — the default of `1` drops the reference
row. `inFilter` narrows the rows further, and accepts either a plain R predicate as
above or PsN's `.gt.` / `.lt.` / `.eq.` string syntax.

These three are alternatives, not combinations: setting more than one is an error
rather than a silent choice by precedence.

---

## Choosing the schedule

The PsN documentation recommends the default five-iteration schedule
`nSamples = c(1000, 1000, 1000, 2000, 2000)` /
`nResample = c(200, 400, 500, 1000, 1000)`.  The rule of thumb is that each
iteration's sample-to-resample ratio should be approximately 5 : 1.  The final
resampled covariance should be stable across iterations — if it is still changing
noticeably between the last two iterations, add more iterations or increase sample
counts.

For parameters that are far from normal (e.g. highly constrained variance components),
setting `thetaInflation` or `omegaInflation` to 1.5–2 widens the initial proposal and
reduces the risk of importance-weight collapse.

---

## References

Dosne, A.-G., Bergstrand, M., & Karlsson, M.O. (2013). An automated sampling
importance resampling procedure for estimating parameter uncertainty. *PAGE 22*,
Abstract 2907.

Dosne, A.-G., Bergstrand, M., Harling, K., & Karlsson, M.O. (2016). Improving the
estimation of parameter uncertainty distributions in nonlinear mixed effects models
using sampling importance resampling. *Journal of Pharmacokinetics and
Pharmacodynamics*, 43(6), 583–596.

Lindbom, L., Ribbing, J., & Jonsson, E.N. (2004). Perl-speaks-NONMEM (PsN) — a Perl
module for NONMEM related programming. *Computer Methods and Programs in Biomedicine*,
75, 85–94.
