---
title: "SIR as the covariance step with setCov()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIR as the covariance step with setCov()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Why a SIR covariance step

An `nlmixr2` fit reports parameter uncertainty from its covariance step: `fit$cov`,
and the standard errors, %RSE and confidence intervals in `fit$parFixedDf` derived
from it. That covariance is an asymptotic approximation. Sampling importance
resampling (SIR) instead samples the uncertainty distribution directly, and its
empirical covariance often describes variance components and correlated
parameters better.

`nlmixr2est::setCov()` switches a fit between covariance methods, such as `"r,s"`,
`"r"`, `"s"` and `"analytic"`. `nlmixr2sir` adds `"sir"` to that list, so SIR can
become *the* covariance of the fit:

```r
setCov(fit, "sir")
```

The standard errors, %RSE and intervals then come from SIR. The covariance the fit
had before is kept in `fit$covList`, so `setCov()` can swap back at any time without
recomputing anything.

For the SIR diagnostics, plots and saved, resumable runs, use `runSIR()`; see
`vignette("runSIR", package = "nlmixr2sir")`. A finished `runSIR()` result can be
installed the same way, as shown [below](#installing-a-runsir-result).

```{r packages, message = FALSE}
library(nlmixr2est)
library(nlmixr2sir)
```

## Fit the model

The same one-compartment theophylline model as in the `runSIR()` vignette:

```{r fit, message = FALSE, results = "hide"}
three_eta_one_cmt <- function() {
  ini({
    tka <- 0.45
    tcl <- 1.00
    tv <- 3.45
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}

fit <- nlmixr2utils::nlmixr2(
  three_eta_one_cmt,
  data = nlmixr2data::theo_sd,
  est = "focei",
  control = list(print = 0L)
)
```

The fit has the covariance its estimation computed:

```{r fit-cov}
estCov <- fit$covMethod
estCov
fit$parFixedDf[, c("Estimate", "SE", "%RSE")]
```

## Install a SIR covariance

`setCov(fit, "sir")` runs SIR at the fit's estimates and installs the result.
Its options live in `sirControl()`. The defaults are the PsN schedule
(7 000 model evaluations over five iterations), which is what you want for a real
analysis. To keep this vignette fast, a two-iteration schedule is used here:

```{r sir-control}
ctl <- sirControl(
  nSamples = c(100, 100),
  nResample = c(20, 40),
  workers = 1L
)
```

```{r setcov-sir, message = FALSE, warning = FALSE}
setCov(fit, "sir", control = ctl)
```

The fit now reports SIR's uncertainty, and the covariance it had before is in
`fit$covList`:

```{r after-sir}
fit$covMethod
names(fit$covList)
fit$parFixedDf[, c("Estimate", "SE", "%RSE")]
```

The SIR result behind the covariance is kept on the fit as `fit$sir`. It is the
same kind of object `runSIR()` returns, so it prints, summarises and plots the same
way:

```{r fit-sir}
class(fit$sir)
attr(fit$sir, "iterationSummary")[, c("iter", "nSamples", "nResample", "nResampled")]
```

## Swapping between covariances

Every covariance a fit has computed stays available. Swapping back to the
estimation's covariance, and then to SIR again, costs nothing: both come from the
cache, so SIR is not rerun.

```{r swap, message = FALSE}
setCov(fit, estCov)
fit$covMethod
setCov(fit, "sir", control = ctl)
fit$covMethod
```

## What the cache is keyed on

SIR starts from a *seed* covariance. Its first proposal is built from it, so a
different seed can give a different answer. A cached SIR covariance is therefore
reused only when both of these are unchanged:

* the `sirControl()` options that shape the result: schedule, inflation, caps,
  recentering, Box-Cox, the random seed and so on. `workers` and `rxThreads` only
  change how fast it runs, so they are not part of the key.
* the seed covariance.

By default the seed is the covariance installed when `setCov(fit, "sir")` is
called. Asking for the same SIR covariance again is therefore not a new
computation:

```{r same, error = TRUE}
setCov(fit, "sir", control = ctl)
```

Installing a different covariance changes the default seed, and the next
`setCov(fit, "sir")` recomputes SIR from it rather than reinstalling a SIR
covariance built from a different starting point:

```{r reseed, message = FALSE, warning = FALSE}
setCov(fit, "r,s")
setCov(fit, "sir", control = ctl)
fit$env$covOptions$sir$seedMethod
```

When `"sir"` is already installed and its options change, the seed it was computed
from is used again. SIR is never seeded from its own result.

Only one SIR covariance is kept at a time: computing a new one replaces the old
one, and `fit$sir` with it.

### Choosing the seed

`sirControl(seedCov =)` picks the seed explicitly: the name of another covariance
on the fit, or a named matrix.

```{r seedcov, eval = FALSE}
setCov(fit, "sir", control = sirControl(seedCov = "r,s"))
setCov(fit, "sir", control = sirControl(seedCov = myCovarianceMatrix))
```

A matrix may be named with `fit$cov` names or SIR names. If it leaves out a block,
such as OMEGA, that block is filled in the same way as a `runSIR()` proposal from a
THETA-only covariance.

## Installing a runSIR() result

A result from `runSIR()` can be installed as the fit's covariance directly:

```{r runsir-install, eval = FALSE}
sir <- runSIR(fit, directory = "theo_sir")
setCov(fit) <- sir
```

This checks that the run was made on this fit (same model, data and estimates),
installs its covariance as `"sir"`, and stores the result as `fit$sir`.

`runSIR()` also registers its covariance in `fit$covList` when it finishes.
`runSIR()` seeds each iteration from its own saved state, so its result is not
the covariance `setCov(fit, "sir")` would compute. A plain `setCov(fit, "sir")`
therefore recomputes rather than reusing it. To swap back to the `runSIR()`
covariance later, install the stored result again:

```{r reinstall, eval = FALSE}
setCov(fit) <- fit$sir
```

## A fit without a covariance step

If the covariance step failed or was switched off, there is no covariance to seed
SIR from:

```{r nocov-fit, message = FALSE, results = "hide"}
fit_nocov <- nlmixr2utils::nlmixr2(
  three_eta_one_cmt,
  data = nlmixr2data::theo_sd,
  est = "focei",
  control = list(print = 0L, covMethod = "")
)
```

`setCov(fit_nocov, "sir")` then seeds SIR from an assumed relative standard error,
30% by default, just as `runSIRControl(rseTheta = 30)` does in the
[`runSIR()` vignette](runSIR.html#running-without-a-covariance-step). It says so,
because that starting width is an assumption rather than an estimate. SIR's
iterations then move the proposal towards the actual uncertainty.

```{r nocov-sir, warning = FALSE}
setCov(fit_nocov, "sir", control = ctl)
fit_nocov$covMethod
fit_nocov$parFixedDf[, c("Estimate", "SE", "%RSE")]
```

`sirControl(rseTheta =, rseOmega =, rseSigma =)` change the assumed widths. They
are only used when the fit has no covariance. The RSE already states the width, so
they cannot be combined with inflation.

## When to use runSIR() instead

`setCov(fit, "sir")` is for making SIR the uncertainty a fit reports. It runs in
memory and writes nothing. Use `runSIR()` when you also want:

* the output directory, with raw results and covariance files;
* recovery of an interrupted run, and `addIterations` to extend a finished one;
* the other proposal sources: `covmatInput` files and `rawresInput`.

Either way, the result can end up as the fit's covariance.
